```{r}
#| child: aaa.Rmd
#| include: false
```

`r descr_models("db_clust", "hdbscan")`

## Tuning Parameters

```{r}
#| label: hdbscan-param-info
#| echo: false
defaults <-
  tibble::tibble(
    tidyclust = c("min_points"),
    default = c("no default")
  )

param <-
  db_clust() |>
  set_engine("hdbscan") |>
  set_mode("partition") |>
  make_parameter_list(defaults)
```

This model has `r nrow(param)` tuning parameters:

```{r}
#| label: hdbscan-param-list
#| echo: false
#| results: asis
param$item
```

The `hdbscan` engine also accepts the engine-specific argument
`min_cluster_size` (passed via `set_engine("hdbscan", min_cluster_size = ...)`).
When supplied, it overrides `min_points` as the value of `minPts` passed to
`dbscan::hdbscan()`. If not supplied, `min_points` is used.

## Translation from tidyclust to the original package (partition)

```{r}
#| label: hdbscan-cls
db_clust(min_points = 5) |>
  set_engine("hdbscan") |>
  set_mode("partition") |>
  translate_tidyclust()
```

## Preprocessing requirements

```{r}
#| child: template-makes-dummies.Rmd
```

```{r}
#| child: template-same-scale.Rmd
```

## What does it mean to predict?

To predict the cluster assignment for a new observation, the nearest training
observation that was not classified as noise is found. The new observation is
assigned to that neighbor's cluster if the distance is at most the neighbor's
core distance; otherwise the new observation is marked as an outlier.

## References

- Campello, R. J. G. B., Moulavi, D., & Sander, J. (2013). Density-Based
  Clustering Based on Hierarchical Density Estimates. In _Advances in Knowledge
  Discovery and Data Mining_ (Vol. 7819, pp. 160–172). Springer.
  https://doi.org/10.1007/978-3-642-37456-2_14

- Campello, R. J. G. B., Moulavi, D., Zimek, A., & Sander, J. (2015).
  Hierarchical density estimates for data clustering, visualization, and
  outlier detection. _ACM Transactions on Knowledge Discovery from Data_,
  10(1), 1–51. https://doi.org/10.1145/2733381

- Hahsler, M., Piekenbrock, M., & Doran, D. (2019). dbscan: Fast Density-Based
  Clustering with R. _Journal of Statistical Software_, 91(1).
  https://doi.org/10.18637/jss.v091.i01
