```{r}
#| child: aaa.Rmd
#| include: false
```

`r descr_models("k_means", "klaR")`

## Tuning Parameters

```{r}
#| label: klaR-param-info
#| echo: false
defaults <- 
  tibble::tibble(tidyclust = c("num_clusters"),
                 default = c("no default"))

param <-
 k_means() |> 
  set_engine("klaR") |> 
  set_mode("partition") |> 
  make_parameter_list(defaults)
```

This model has `r nrow(param)` tuning parameters:

```{r}
#| label: klaR-param-list
#| echo: false
#| results: asis
param$item
```

## Translation from tidyclust to the original package (partition)

```{r}
#| label: klaR-cls
k_means(num_clusters = integer(1)) |> 
  set_engine("klaR") |> 
  set_mode("partition") |> 
  translate_tidyclust()
```

## Preprocessing requirements

Only categorical variables are accepted, along with numerics with few unique values.

## References

- Huang, Z. (1997) A Fast Clustering Algorithm to Cluster Very Large Categorical Data Sets in Data Mining. in KDD: Techniques and Applications (H. Lu, H. Motoda and H. Luu, Eds.), pp. 21-34, World Scientific, Singapore.

- MacQueen, J. (1967) Some methods for classification and analysis of multivariate observations. In Proceedings of the Fifth Berkeley Symposium on Mathematical Statistics and Probability, eds L. M. Le Cam & J. Neyman, 1, pp. 281-297. Berkeley, CA: University of California Press.
