```{r}
#| child: aaa.Rmd
#| include: false
```

`r descr_models("k_means", "ClusterR")`

## Tuning Parameters

```{r}
#| label: ClusterR-param-info
#| echo: false
defaults <- 
  tibble::tibble(tidyclust = c("num_clusters"),
                 default = c("no default"))

param <-
 k_means() |> 
  set_engine("ClusterR") |> 
  set_mode("partition") |> 
  make_parameter_list(defaults)
```

This model has `r nrow(param)` tuning parameters:

```{r}
#| label: ClusterR-param-list
#| echo: false
#| results: asis
param$item
```

## Translation from tidyclust to the original package (partition)

```{r}
#| label: ClusterR-cls
k_means(num_clusters = integer(1)) |> 
  set_engine("ClusterR") |> 
  set_mode("partition") |> 
  translate_tidyclust()
```

## Preprocessing requirements

```{r}
#| child: template-makes-dummies.Rmd
```

```{r}
#| child: template-same-scale.Rmd
```

## References

- Forgy, E. W. (1965). Cluster analysis of multivariate data: efficiency vs interpretability of classifications. Biometrics, 21, 768–769.

- Hartigan, J. A. and Wong, M. A. (1979). Algorithm AS 136: A K-means clustering algorithm. Applied Statistics, 28, 100–108. doi:10.2307/2346830.

- Lloyd, S. P. (1957, 1982). Least squares quantization in PCM. Technical Note, Bell Laboratories. Published in 1982 in IEEE Transactions on Information Theory, 28, 128–137.

- MacQueen, J. (1967). Some methods for classification and analysis of multivariate observations. In Proceedings of the Fifth Berkeley Symposium on Mathematical Statistics and Probability, eds L. M. Le Cam & J. Neyman, 1, pp. 281–297. Berkeley, CA: University of California Press.
