```{r}
#| child: aaa.Rmd
#| include: false
```

`r descr_models("k_means", "clustMixType")`

## Tuning Parameters

```{r}
#| label: clustMixType-param-info
#| echo: false
defaults <- 
  tibble::tibble(tidyclust = c("num_clusters"),
                 default = c("no default"))

param <-
 k_means() |> 
  set_engine("clustMixType") |> 
  set_mode("partition") |> 
  make_parameter_list(defaults)
```

This model has `r nrow(param)` tuning parameters:

```{r}
#| label: clustMixType-param-list
#| echo: false
#| results: asis
param$item
```

## Translation from tidyclust to the original package (partition)

```{r}
#| label: clustMixType-cls
k_means(num_clusters = integer(1)) |> 
  set_engine("clustMixType") |> 
  set_mode("partition") |> 
  translate_tidyclust()
```

## Preprocessing requirements

Both categorical and numeric predictors are required.

## References

- Szepannek, G. (2018): clustMixType: User-Friendly Clustering of Mixed-Type Data in R, The R Journal 10/2, 200-208, doi:10.32614/RJ-2018-048.

- Aschenbruck, R., Szepannek, G., Wilhelm, A. (2022): Imputation Strategies for Clustering Mixed‑Type Data with Missing Values, Journal of Classification, doi:10.1007/s00357-022-09422-y.

- Z.Huang (1998): Extensions to the k-Means Algorithm for Clustering Large Data Sets with Categorical Variables, Data Mining and Knowledge Discovery 2, 283-304.
