```{r, child = "aaa.Rmd", include = FALSE}
```

`r descr_models("mean_shift", "LPCM")`

## Tuning Parameters

```{r stats-param-info, echo = FALSE}
defaults <-
  tibble::tibble(
    tidyclust = c("bandwidth"),
    default = c("no default")
  )

param <-
  mean_shift() %>%
  set_engine("LPCM") %>%
  set_mode("partition") %>%
  make_parameter_list(defaults)
```

This model has `r nrow(param)` tuning parameters:

```{r LPCM-param-list, echo = FALSE, results = "asis"}
param$item
```

## Translation from tidyclust to the original package (partition)

```{r stats-cls}
mean_shift(bandwidth = 0.5) %>%
  set_engine("LPCM") %>%
  set_mode("partition") %>%
  translate_tidyclust()
```

## Preprocessing requirements

```{r child = "template-makes-dummies.Rmd"}
```

`LPCM::ms()` scales each variable internally to the unit range before applying
the Gaussian kernel, so the `bandwidth` value lives on the scaled scale rather
than the raw data scale. Bandwidths between roughly `0.05` and `1` are typical;
smaller values find more clusters and larger values merge them.

## What does it mean to predict?

To predict the cluster assignment for a new observation, the mean shift
procedure is run from the new point until it converges to a mode. The
observation is then assigned to the cluster of the nearest discovered training
mode by Euclidean distance.


## References

- Cheng, Y. (1995). Mean shift, mode seeking, and clustering. IEEE Transactions on Pattern Analysis and Machine Intelligence, 17(8), 790–799. https://doi.org/10.1109/34.400568

- Comaniciu, D., & Meer, P. (2002). Mean shift: A robust approach toward feature space analysis. IEEE Transactions on Pattern Analysis and Machine Intelligence, 24(5), 603–619. https://doi.org/10.1109/34.1000236

- Einbeck, J., Evers, L., & Hinchliff, K. (2010). Data compression and regression based on local principal curves. In A. Fink, B. Lausen, W. Seidel, & A. Ultsch (Eds.), Advances in Data Analysis, Data Handling and Business Intelligence (pp. 701–712). Springer.
