---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyclust <img src="man/figures/logo.svg" align="right" height="139" />

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/EmilHvitfeldt/tidyclust/branch/main/graph/badge.svg)](https://app.codecov.io/gh/EmilHvitfeldt/tidyclust?branch=main)
[![R-CMD-check](https://github.com/EmilHvitfeldt/tidyclust/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/EmilHvitfeldt/tidyclust/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of tidyclust is to provide a tidy, unified interface to clustering models. The packages is closely modeled after the [parsnip](https://parsnip.tidymodels.org/) package.

## Installation

You can install the development version of tidyclust from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("EmilHvitfeldt/tidyclust")
```

Please note that this package currently requires a [branch of the workflows](https://github.com/tidymodels/workflows/tree/tidyclust) package to work. Use with caution.

## Example

The first thing you do is to create a `cluster specification`. For this example we are creating a K-means model, using the `stats` engine.

```{r}
library(tidyclust)

kmeans_spec <- k_means(num_clusters = 3) %>%
  set_engine("stats") 

kmeans_spec
```

This specification can then be fit using data.

```{r}
kmeans_spec_fit <- kmeans_spec %>%
  fit(~., data = mtcars)
kmeans_spec_fit
```

Once you have a fitted tidyclust object, you can do a number of things. `predict()` returns the cluster a new observation belongs to

```{r}
predict(kmeans_spec_fit, mtcars[1:4, ])
```

`extract_cluster_assignment()` returns the cluster assignments of the training observations

```{r}
extract_cluster_assignment(kmeans_spec_fit)
```

and `extract_clusters()` returns the locations of the clusters

```{r}
extract_centroids(kmeans_spec_fit)
```

