---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# celery

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/EmilHvitfeldt/celery/branch/main/graph/badge.svg)](https://app.codecov.io/gh/EmilHvitfeldt/celery?branch=main)
[![R-CMD-check](https://github.com/EmilHvitfeldt/celery/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/EmilHvitfeldt/celery/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of celery is to provide a tidy, unified interface to clustering models. The packages is closely modeled after the [parsnip](https://parsnip.tidymodels.org/) package.

## Installation

You can install the development version of celery from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("EmilHvitfeldt/celery")
```

## Example

The first thing you do is to create a `cluster specification`. For this example we are creating a K-means model, using the `stats` engine.

```{r}
library(celery)

k_means_spec <- k_means(k = 3) %>%
  set_engine_celery("stats") 

k_means_spec
```

This specification can then be fit using data.

```{r}
k_means_spec %>%
  fit(~., data = mtcars)
```
