---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyclust <img src="man/figures/logo.svg" align="right" height="139" />

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/tidymodels/tidyclust/branch/main/graph/badge.svg)](https://app.codecov.io/gh/tidymodels/tidyclust?branch=main)
[![R-CMD-check](https://github.com/tidymodels/tidyclust/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tidymodels/tidyclust/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of tidyclust is to provide a tidy, unified interface to clustering models. The packages is closely modeled after the [parsnip](https://parsnip.tidymodels.org/) package.

## Installation

You can install the released version of tidyclust from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("tidyclust")
```


and the development version of tidyclust from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("tidymodels/tidyclust")
```

## Example

The first thing you do is to create a `cluster specification`. For this example we are creating a K-means model, using the `stats` engine.

```{r}
library(tidyclust)
set.seed(1234)

kmeans_spec <- k_means(num_clusters = 3) |>
  set_engine("stats")

kmeans_spec
```

This specification can then be fit using data.

```{r}
kmeans_spec_fit <- kmeans_spec |>
  fit(~., data = mtcars)
kmeans_spec_fit
```

Once you have a fitted tidyclust object, you can do a number of things. `predict()` returns the cluster a new observation belongs to

```{r}
predict(kmeans_spec_fit, mtcars[1:4, ])
```

`extract_cluster_assignment()` returns the cluster assignments of the training observations

```{r}
extract_cluster_assignment(kmeans_spec_fit)
```

and `extract_centroids()` returns the locations of the clusters

```{r}
extract_centroids(kmeans_spec_fit)
```

## Visual comparison of clustering methods

Below is a visualization of the available models and how they compare using 2 dimensional toy data sets.

```{r}
#| label: comparison
#| echo: false
#| message: false
#| fig-asp: !expr 1/2
#| fig-dpi: 105
#| fig-format: svglite
#| fig-alt: "Mock comparison for different clustering methods for different data sets. Each row correspods to a clustering method, each column corresponds to a data set type."
library(tidymodels)
library(tidyclust)
set.seed(1234)

make_circles <- function(n) {
  x <- seq(0, pi * 2, length.out = n)
  x <- cos(x) * c(0.5, 1)
  x <- x + rnorm(n, sd = 0.05)

  y <- seq(0, pi * 2, length.out = n)
  y <- sin(y) * c(0.5, 1)
  y <- y + rnorm(n, sd = 0.05)

  out <- data.frame(x, y)
  attr(out, "name") <- "circles"
  out
}

make_halves <- function(n) {
  x <- seq(0, pi * 2, length.out = n)
  x <- cos(x) + rep(c(0, 1), each = n / 2)
  x <- x + rnorm(n, sd = 0.05)

  y <- seq(0, pi * 2, length.out = n)
  y <- sin(y) + rep(c(0, 0.5), each = n / 2)
  y <- y + rnorm(n, sd = 0.05)
  y <- y - 0.25
  y <- y * (1 / 0.75)

  out <- data.frame(x, y)
  attr(out, "name") <- "halves"
  out
}

make_uniform <- function(n) {
  x <- runif(n, min = -1, max = 1)
  y <- runif(n, min = -1, max = 1)

  out <- data.frame(x, y)
  attr(out, "name") <- "uniform"
  out
}

make_blobs <- function(n) {
  x <- rep(c(1, 2, 3), length.out = n)
  x <- x + rnorm(n, sd = 0.5)
  x <- (x - min(x)) / (max(x) - min(x)) * 2 - 1

  y <- rep(c(1, 4, 2), length.out = n)
  y <- y + rnorm(n, sd = 0.5)
  y <- (y - min(y)) / (max(y) - min(y)) * 2 - 1

  out <- data.frame(x, y)
  attr(out, "name") <- "blobs"
  out
}

augment_model <- function(model, data) {
  model |>
    fit(~., data = data) |>
    augment(new_data = data) |>
    mutate(
      data_name = attr(data, "name"),
      model_name = class(model)[1]
    )
}

circle_data <- make_circles(500)
halves_data <- make_halves(500)
uniform_data <- make_uniform(500)
blobs_data <- make_blobs(500)

colors <- c("#E49E68", "#6899E4", "#E068E4")

expand_grid(
  models = list(
    k_means(num_clusters = 3),
    hier_clust(num_clusters = 3)
  ),
  datasets = list(
    circle_data,
    halves_data,
    blobs_data,
    uniform_data
  )
) |>
  pmap_dfr(\(.x) augment_model(.x, .y)) |>
  ggplot(aes(x, y, color = .pred_cluster)) +
  geom_point(alpha = 0.5) +
  facet_grid(model_name ~ data_name, scales = "free", switch = "y") +
  scale_color_manual(values = colors) +
  theme_void() +
  theme(
    strip.text.x = element_blank(),
    strip.text.y = element_text(size = 15)
  ) +
  guides(color = "none")
```

## Contributing

This project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

- For questions and discussions about tidymodels packages, modeling, and machine learning, please [post on RStudio Community](https://forum.posit.co/new-topic?category_id=15&tags=tidymodels,question).

- If you think you have encountered a bug, please [submit an issue](https://github.com/tidymodels/tidyclust/issues).

- Either way, learn how to create and share a [reprex](https://reprex.tidyverse.org/articles/articles/learn-reprex.html) (a minimal, reproducible example), to clearly communicate about your code.

- Check out further details on [contributing guidelines for tidymodels packages](https://www.tidymodels.org/contribute/) and [how to get help](https://www.tidymodels.org/help/).
Footer
