---
title: "k-means Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k-means Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(celery)
library(palmerpenguins)
library(tidymodels)
```

## Fit

```{r}
kmeans_spec <- k_means(k = 5) %>%
  set_engine_celery("stats")

penguins_rec_1 <- recipe(~ ., data = penguins) %>%
  update_role(species, island, new_role = "demographic") %>%
  step_dummy(sex)
  
  
penguins_rec_2 <- recipe(species ~ ., data = penguins) %>%
  step_dummy(sex, island)

wflow_1 <- workflow() %>%
  add_model(kmeans_spec) %>%
  add_recipe(penguins_rec_1)


wflow_2 <- workflow() %>%
  add_model(kmeans_spec) %>%
  add_recipe(penguins_rec_2)
```

We need workflows!

```{r}
# dropping NA first so rows match up later, this is clunky
pen_sub <- penguins %>% 
  drop_na() %>%
  select(-species, -island, -sex)

kmeans_fit <- kmeans_spec %>% fit( ~., pen_sub)

kmeans_fit %>%
  predict(new_data = pen_sub)
```

* Needed flexclust install; probably not necessary, we could implement for kmeans with just dists

* Missing values should probably return an NA prediction.  Or for k-means, imputation isn't crazy...

* We want a consistent return, and k-means is randomized.  We should agree to a consistent default ordering throughout celery - maybe by size (number of members) or something?


```{r}
penguins %>%
  drop_na() %>%
  mutate(
    preds = predict(kmeans_fit, new_data = pen_sub)$.pred_cluster
  ) %>%
  count(preds, sex, species)
```

## Diagnostics

* Measure cluster enrichment with demographic variables (an official recipe designation?).  Include Chi-Square or ANOVA tests???

* Exploit confusion matrix from `yardstick`

* Characterization of clusters, presumably by centers?

* Automatic plot?!

