---
title: "Frequent Itemset Mining"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequent Itemset Mining}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

```{r}
library(workflows)
library(parsnip)
```

Load libraries:

```{r setup}
library(tidyclust)
library(arules)
set.seed(838383)
```

Load and clean a dataset:

```{r}
data(Groceries)

# convert to data frame
groceries <- as.data.frame(as(Groceries, "matrix")) %>%
  dplyr::mutate(across(everything(), ~.*1))
```

## A Brief Introduction to Frequent Itemset Mining

*Frequent Itemset Mining* (FIM) is a fundamental technique in data mining that
identifies sets of items that frequently appear together in
transactional datasets. These itemsets are often used to uncover
meaningful patterns, such as associations between items, which can then
be leveraged to generate *association rules*.

For example, in a supermarket transaction database, frequent itemset
mining can identify groups of products that are commonly purchased
together, such as `{milk, bread, eggs}`. These insights are valuable for
applications like recommendation systems, inventory management, and
targeted marketing.

The key to frequent itemset mining is determining the sets of items that
satisfy a user-defined threshold called the **minimum support**, where
support is defined as the proportion of transactions in which a
particular itemset appears.

### Methods of Frequent Itemset Mining

Efficiently discovering these frequent itemsets is a computational
challenge, and several algorithms have been developed to address this
challenge. The two implemented in `{tidyclust}` are the **Apriori**
algorithm and the **Eclat** algorithm.

#### Finding Frequent Itemsets with the Apriori Algorithm

The *Apriori* algorithm is one of the earliest and most widely known
methods for frequent itemset mining. It is based on the **Apriori
Principle** (also known as **Downward Closure Property**): any subset of
a frequent itemset must also be frequent.

#### Process of the Apriori Algorithm

1.  **Initialization**: Begin by identifying all individual items
    (1-itemsets) that satisfy the minimum support threshold. These are
    called *frequent 1-itemsets*.

2.  **Candidate Generation**: Use the frequent itemsets from the
    previous step to generate candidate itemsets of the next size (e.g.
    combine frequent 1-itemsets to create candidate 2-itemsets).

3.  **Prune Candidates**: Eliminate candidate itemsets that have subsets
    not found to be frequent.

4.  **Support Counting**: Scan the dataset to count the occurrences of
    each candidate itemset.

5.  **Iteration**: Repeat steps 2–4 for larger itemsets until no more
    frequent itemsets can be generated.

[![Apriori Princple
Example](images/clipboard-4074102307.png){width="650"}](https://chih-ling-hsu.github.io/2017/03/25/apriori)

The Apriori algorithm is computationally expensive due to repeated
database scans and the generation of numerous candidates. However, its
pruning strategy significantly reduces the search space compared to a
naïve approach.

[Source](https://dl.acm.org/doi/pdf/10.1145/170036.170072)

#### Finding Frequent Itemsets with the Eclat Algorithm

The *Eclat* (Equivalence Class Transformation) algorithm is an
alternative to Apriori that uses a depth-first search strategy and
vertical data representation. Instead of scanning the dataset
repeatedly, Eclat represents transactions as *tid-lists* (transaction ID
lists), which map each item or itemset to the IDs of transactions in
which it appears.

#### Process of the Eclat Algorithm

1.  **Vertical Data Representation**: Transform the dataset into a
    vertical format, where each item is associated with a list of
    transaction IDs.

2.  **Intersect Tid-lists**: Generate frequent itemsets by recursively
    intersecting the tid-lists of individual items to form larger
    itemsets. The intersection results in a new tid-list, representing
    the transactions containing the larger itemset.

3.  **Check Support**: The length of the resulting tid-list determines
    the support of the itemset. Remove itemsets not found to be
    frequent.

4.  **Recursive Search**: Continue the process for all itemsets until no
    further frequent itemsets can be found.

[![Bookstore
database](images/clipboard-2860408154.png){width="325"}](https://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=846291)

[![Computing support of itemsets via tid-list
intersections](images/clipboard-3198084587.png){width="650"}](https://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=846291)

Eclat is generally more efficient than Apriori for datasets with many
transactions but fewer unique items, as it avoids the need for multiple
scans of the dataset. However, its performance can degrade for datasets
with very large tid-lists.

[Source](https://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=846291)

## **`freq_itemsets` specification in {tidyclust}**

To specify a frequent itemsets mining model in `tidyclust`, simply
choose a value of `min_support` and (optionally) a mining method:

```{r}
fi_spec <- freq_itemsets(
  min_support = 0.05,
  mining_method = "eclat"
  ) %>%
  set_engine("arules") %>%
  set_mode("partition")

fi_spec
```

Currently, the only supported engine is `arules`. The default mining 
method is eclat because it is generally faster in most practical cases. A 
default `min_support` value is not provided as it varies significantly depending 
on the data characteristics.

## **Fitting `freq_itemsets` models**

We fit the model to the data in the usual way:

```{r}
fi_fit <- fi_spec %>%
  fit(~ .,
    data = groceries
  )

fi_fit %>%
  summary()
```

We can not extract the standard `tidyclust` summary list since centroids are not 
useful for frequent itemsets, however we can extract the frequent itemsets:

```{r}
arules::inspect(fi_fit$fit)
```

Note that, although the frequent itemset algorithm is not focused on cluster's
like other unsupervised learning algorithms, we have created clusters based on
the itemsets. For each item, we find all itemsets that include that item:

- Itemsets with the largest size are selected as the "dominate" itemset for the 
item. If there is a tie in size, the itemset with the highest support is selected. 
This prioritization aligns with findings that larger itemsets with higher 
support exhibit greater predictive utility.

- If an item has already been assigned a cluster, the algorithm compares the 
current "best" itemset with the itemset under consideration and re-prioritizes 
based on size and support. This process repeats until no items are reassigned to 
a new cluster (convergence).

- Items that appear in no frequent itemsets are labeled as outliers 
(Cluster_0_X) while items within frequent itemsets are assigned sequential 
cluster IDs (Cluster_1, Cluster_2, etc.). 

```{r}
fi_fit %>% 
  extract_cluster_assignment()
```

## Prediction

Since frequent itemset mining identifies patterns in co-occurring items rather 
than learning a predictive function, the notion of "prediction" is not as 
straightforward as in supervised learning. However, given a set of frequent 
itemsets from historical data, it is possible to estimate the likelihood that a 
missing item in new data is present based on observed co-occurring items.

The `predict()` function utilizes frequent itemsets and their support values to 
estimate probabilities for missing items in new transactions. For each row in 
`new_data`, the function identifies observed items and missing items. It then 
searches for frequent itemsets that contain both the missing item and at least 
one observed item. Using the support values of these itemsets, it estimates the 
probability that the missing item is present based on the confidence of 
association between observed and missing items. If no relevant itemsets are 
found, the item's global support (frequency in training data) is used as a 
fallback prediction.

The function fills in missing values with these probability estimates, 
effectively "predicting" the likelihood of item presence based on historical 
co-occurrence patterns. The type argument allows returning raw prediction 
probabilities ('raw') or binary predictions based on a 0.5 threshold ('cluster').

We display the predicted values in the column `.pred_item`, and the observed 
values in the column `.obs_item`.

```{r}
new_data <- groceries[1:5,]  %>%
  tidyclust:::random_na_with_truth(na_prob = 0.3)

results <- fi_fit %>%
  predict(new_data$na_data)

results$.pred_cluster[[1]]
```

The function `random_na_with_truth()` is used for testing purposes to randomly 
assign values to be `NA`, however it can be accessed using in `tidyclust` using 
the `:::`.

Additionally, we can extract the nested predicted output to be formatted in a 
single data frame, filling in the `NA` values with their predicted value using
`extract_itemset_predictions()`.

```{r}
results %>%
  extract_itemset_predictions
```

## Evaluation Metrics

While support values traditionally assess frequent itemset quality, they do not 
guarantee predictive performance. Since the `predict()` methodology resembles a 
recommender system, the results should be evaluated using similar metrics. 
Common metrics such as root mean squared error (RMSE), accuracy, precision, and 
recall are implemented in the `yardstick` package.

To prepare the `predict()` output for metric calculation, we provide a new 
function `augment_itemset_predict()`.

```{r}
# Generate data to predict on
na_result <- tidyclust:::random_na_with_truth(groceries[1:5,], na_prob = 0.3)
new_data <- na_result$na_data # In a real scenario, this would be new, untrained on, data
truth_output <-  na_result$truth # In a real scenario, this would be a separate holdout set

# Example for RMSE (using type = 'raw')
fi_fit %>%
  predict(new_data = new_data, type = 'raw') %>%
  augment_itemset_predict(truth_output = truth_output) %>%
  yardstick::rmse(truth, preds)

# Example for Precision (using type = 'cluster')
fi_fit %>%
  predict(new_data = new_data, type = 'cluster') %>%
  augment_itemset_predict(truth_output = truth_output) %>%
  dplyr::mutate(
    truth = factor(truth, levels = c(0, 1)),
    preds = factor(preds, levels = c(0, 1))
  ) %>%
  yardstick::precision(truth, preds)
```

When using RMSE, the raw `predict()` output should be used, while accuracy, 
precision, and recall will use the cluster output or user thresholded raw 
output. Caution should be used when looking at accuracy, precision, and recall 
for imbalanced datasets (where items are infrequently purchased). In such cases, 
F1-Score offers a balance between precision and recall, and precision-recall 
(PR) curves aid in determining the best threshold value.

```{r}
fi_fit %>%
  predict(new_data = new_data, type = 'raw') %>%
  augment_itemset_predict(truth = truth_output) %>%
  dplyr::mutate(truth = factor(truth, levels = c(0, 1))) %>%
  yardstick::pr_curve(truth, preds) %>%
  autoplot()
```

Each point on the PR curve represents the precision and recall of the model at a 
specific threshold. By varying this threshold, different precision and recall 
values are obtained, creating the curve. The ideal curve is close to the 
top-right corner, indicating high precision and high recall.

## Hyperparameter Tuning

The sole parameter capable of being tuned in a FIM model is `min_support.` 
Selecting the correct value is imperative for finding useful frequent itemsets. 
The default grid for `min_support` is from 0.1 to 0.5. The lower bound of 0.1 was 
chosen to avoid reporting too many frequent itemsets, even for smaller datasets, 
while the upper bound of 0.5 was selected since it ensures that each frequent 
itemset has a support of at least 50%.

```{r}
dials::grid_regular(
  dials::min_support(),
  levels = 10
)
```

Usually, the above object is paired with `tune_cluster`, however cross-validation 
is not currently implemented for FIM. Future work will focus on improving tuning 
and implementing cross-validation.
