---
title: "Hierarchical Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Setup

```{r}
library(workflows)
library(parsnip)
```

Load libraries:

```{r}
#| label: setup
library(tidyclust)
library(tidyverse)
library(tidymodels)
set.seed(838383)
```

Load and clean a dataset:

```{r}
data("penguins", package = "modeldata")

penguins <- penguins |>
  select(bill_length_mm, bill_depth_mm) |>
  drop_na()

# shuffle rows
penguins <- penguins |>
  sample_n(nrow(penguins))
```

If you have not yet read the `k_means` vignette, we recommend reading that first;
functions that are used in this vignette are explained in more detail there.



## A brief introduction to hierarchical clustering

*Hierarchical Clustering*, sometimes called *Agglomerative Clustering*, is a 
method of *unsupervised* learning that produces a *dendrogram*, which can be used
to partition observations into clusters.  

The hierarchical clustering process begins with each observation in it's own
cluster; i.e., *n* clusters for *n* observations.

```{r}
#| echo: false
#| fig-alt: "scatter chart. 5 circles are randomly located, and labeled a, b, c, d, and e."
fake_dat <- tibble(
  x = sort(runif(5)),
  y = runif(5),
  lab = letters[1:5]
)

fake_dat |>
  ggplot(aes(x, y)) +
  geom_point(shape = fake_dat$lab, size = 4) +
  geom_point(shape = 1, size = 7, stroke = 1, color = "dark grey") +
  theme_minimal() +
  ylim(c(-0.1, 1.1)) +
  xlim(c(-0.1, 1.1))
```

The closest two observations are then joined together into a single cluster.


```{r}
#| echo: false
#| fig-alt: "scatter chart. 5 circles are randomly located, and labeled a, b, c, d, and e. One of the circles are replacing 2 of the previous circles."
fake_dat_2 <- bind_rows(
  fake_dat[-c(1:2), -3],
  summarize_all(fake_dat[1:2, -3], mean)
) |>
  mutate(
    size = c(rep(1, 3), suppressWarnings(dist(fake_dat)[1]))
  )

fake_dat |>
  ggplot(aes(x, y)) +
  geom_point(shape = fake_dat$lab, size = 4) +
  geom_point(
    data = fake_dat_2,
    aes(x = x, y = y),
    shape = 1,
    size = 7 / fake_dat_2$size,
    stroke = 1,
    color = "dark grey"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(c(-0.1, 1.1)) +
  xlim(c(-0.1, 1.1))
```

This process continues, with the closest two clusters being joined (or 
"agglomerated") at each step.


```{r}
#| echo: false
#| fig-alt: "scatter chart. 5 circles are randomly located, and labeled a, b, c, d, and e. One of the circles are replacing 2 of the previous circles."
fake_dat_3 <- bind_rows(
  fake_dat[-c(1:3), -3],
  summarize_all(fake_dat[1:3, -3], mean)
) |>
  mutate(
    size = c(rep(1, 2), 0.09)
  )

p1 <- fake_dat |>
  ggplot(aes(x, y)) +
  geom_point(shape = fake_dat$lab, size = 4) +
  geom_point(
    data = fake_dat_3,
    aes(x = x, y = y),
    shape = 1,
    size = 7 / fake_dat_3$size,
    stroke = 1,
    color = "dark grey"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(c(-0.1, 1.1)) +
  xlim(c(-0.1, 1.1))


fake_dat_4 <- bind_rows(
  summarize_all(fake_dat[1:3, -3], mean),
  summarize_all(fake_dat[4:5, -3], mean),
) |>
  mutate(
    size = c(0.09, 0.09)
  )

p2 <- fake_dat |>
  ggplot(aes(x, y)) +
  geom_point(shape = fake_dat$lab, size = 4) +
  geom_point(
    data = fake_dat_4,
    aes(x = x, y = y),
    shape = 1,
    size = 7 / fake_dat_4$size,
    stroke = 1,
    color = "dark grey"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(c(-0.1, 1.1)) +
  xlim(c(-0.1, 1.1))

library(patchwork)
p1 + p2
```


The result of the process is a **dendrogram**, which shows the joining of clusters
in tree form:

```{r}
#| echo: false
#| fig-alt: "Dendrogram chart. With 5 observations."
hc <- hclust(dist(fake_dat))

plot(hc, labels = fake_dat$lab)
```

### Clusters from dendrogram

To produce a partition-style cluster assignment from the dendrogram, one must
"cut" the tree at a chosen height:

```{r}
#| echo: false
#| fig-alt: "Dendrogram chart. With 5 observations. A horizontal like at 0.6 cuts the dendrogram into 3 clusters."
plot(hc, labels = fake_dat$lab)
abline(h = 0.6, lty = 2, col = "dark grey")
```

The observations that remain joined in the dendrogram below the cut height are
considered to be in a cluster together:

```{r}
#| echo: false
tibble(
  observation = fake_dat$lab,
  cluster_assignment = cutree(hc, h = 0.6)
)
```


### Methods of agglomeration

At every step of the agglomeration, we measure distances between current
clusters.  With each cluster containing (possibly) multiple points, what does
it mean to measure distance?

There are four common approaches to cluster-cluster distancing, aka "linkage":

1. **single linkage:** The distance between two clusters is the distance between
the two **closest** observations.

2. **average linkage:** The distance between two clusters is the average of all
distances between observations in one cluster and observations in the other.

3. **complete linkage:** The distance between two clusters is the distance between
the two **furthest** observations.

4. **centroid method:** The distance between two clusters is the distance between
their centroids (geometric mean or median).

5. **Ward's method:** The distance between two clusters is proportional to the 
increase in **error sum of squares (ESS)** that would result from joining them.
The ESS is computed as the sum of squared distances between observations in a
cluster, and the centroid of the cluster.


It is also worth mentioning the **McQuitty method**, which retains information
about previously joined clusters to measure future linkage distance.  This 
method is currently supported for model fitting, but not for prediction, in
`tidyclust`.


## `hier_clust` specification in {tidyclust}

To specify a hierarchical clustering model in `tidyclust`, 
simply choose a value of `num_clusters` and (optionally) a linkage method:

```{r}
hc_spec <- hier_clust(
  num_clusters = 3,
  linkage_method = "average"
)

hc_spec
```

Currently, the only supported engine is `stats::hclust()`.  The default linkage
<!-- method of this method is *complete linkage*. -->

<!-- Note that the hierarchical clustering algorithm does not need a number of clusters -->
<!-- specified to perform the aggomeration and produce a dendrogram, only to determine -->
<!-- a cut of the tree, so this parameter -->
<!-- does not actually need to be specified in advance. -->


<!-- ```{r} -->
<!-- hc_spec_blank <- hier_clust()  -->

<!-- hc_spec_blank -->
<!-- ``` -->


## Fitting `hier_clust` models

We fit the model to data in the usual way:

```{r}
hc_fit <- hc_spec |>
  fit(~ bill_length_mm + bill_depth_mm,
    data = penguins
  )

hc_fit |>
  summary()
```

To produce a dendrogram plot, access the engine fit:
(Although as we see below, dendrograms are often not very informative for moderate
to large size datasets.)

```{r}
#| fig-alt: "Dendrogram chart. With too many observations to be able to clearly see anything"
hc_fit$fit |> plot()
```

We can also extract the standard `tidyclust` summary list:

```{r}
hc_summary <- hc_fit |> extract_fit_summary()

hc_summary |> str()
```

Note that, although the hierarchical clustering algorithm is not focused on 
cluster centroids in the same way $k$-means is, we are still able to compute
the geometric mean over the predictors for each cluster:

```{r}
hc_fit |> extract_centroids()
```


## Prediction

To predict the cluster assignment for a new observation, we find the closest
cluster.  How we measure "closeness" is dependent on the specified type of
linkage in the model:

* *single linkage*:  The new observation is assigned to the same cluster as its 
nearest observation from the training data.

* *complete linkage*:  The new observation is assigned to the cluster with the
smallest maximum distances between training observations and the new observation.

* *average linkage*: The new observation is assigned to the cluster with the
smallest average distances between training observations and the new observation.

* *centroid method:* The new observation is assigned to the cluster with the
closest centroid, as in prediction for `k_means`.

* *Ward's method:* The new observation is assigned to the cluster with the 
smallest increase in **error sum of squares (ESS)** due to the new addition.
The ESS is computed as the sum of squared distances between observations in a
cluster, and the centroid of the cluster.

```{r}
hc_preds <- hc_fit |> predict(penguins)

hc_preds
```

It's important to note that there is no guarantee that `predict()` on the 
training data will produce the same results as `extract_cluster_assignments()`.
The process by which clusters are created during the agglomerations results in
a particular partition; but if a training observation is treated as new data, it
is predicted in the same manner as truly new information.

```{r}
bind_cols(
  hc_preds,
  extract_cluster_assignment(hc_fit)
)
```


<!-- Overriding cluster partition: -->

<!-- ```{r} -->
<!-- hc_fit |> extract_cluster_assignment(num_clusters = 5) -->
<!-- hc_fit |> extract_cluster_assignment(cut_height = 5) -->
<!-- ``` -->


## Reconciling partitions

Suppose we have produced cluster assignments from two models:  a hierarchical
clustering model with three clusters (as above) and a $k$-means clustering model
 with five clusters (below).  How can we combine these assignments?

```{r}
km_spec <- k_means(num_clusters = 5)
km_fit <- km_spec |>
  fit(~., data = penguins)

km_preds <- predict(km_fit, penguins, prefix = "KM_")
hc_preds <- predict(hc_fit, penguins, prefix = "HC_")
```

We notice that the three-cluster assignments from `hier_clust` do not line up
perfectly with the five-cluster assignments from `k_means`.  

```{r}
tibble(
  hc = hc_preds$.pred_cluster,
  km = km_preds$.pred_cluster
) |>
  count(hc, km)
```

However, they are not fully unrelated assignments.  For example, all of 
`KM_2` in the $k$-means assignment fell inside `HC_1` for the hierarchical
assignments.

Our goal is to relabel the five $k$-means clusters to match the three cluster names
in the hierarchical output.  This can be accomplished with `reconcile_clusterings_mapping()`.

This function expects two vectors of cluster labels as input.  The first is the
label that will be matched, and the second is the label that will be recoded
to the first.

If we are not trying to simply match names across two same-size clusterings,
the option `one_to_one` must be set to `FALSE`.

```{r}
reconcile_clusterings_mapping(
  primary = hc_preds$.pred_cluster,
  alternative = km_preds$.pred_cluster,
  one_to_one = FALSE
)
```

In this example, we can see that `KM_1`, `KM_2`, `KM_5` have been matched to 
`HC_1`; and `KM_3` and `KM_4` have been matched to `HC_2`.  Notice that no
clusters from the `KM` set were matched to `HC_3`; evidently, this is a small
cluster that did not manifest clearly in the $k$-means clustering.
