---
title: "k-means"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{k-means}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Setup

```{r}
library(workflows)
library(parsnip)
```

Load libraries:

```{r}
#| label: setup
library(tidyclust)
library(tidyverse)
library(tidymodels)
```

```{r}
#| label: setup_secret
#| echo: false
library(ggforce)
set.seed(838383)
```

Load and clean a dataset:

```{r}
data("penguins", package = "modeldata")

penguins <- penguins |>
  select(bill_length_mm, bill_depth_mm) |>
  drop_na()


# shuffle rows
penguins <- penguins |>
  sample_n(nrow(penguins))
```

At the end of this vignette, you will find a brief overview of the k-means
algorithm, as well as some further algorithmic variant details, for those who
would like a reference.

## k-means specification in {tidyclust}

To specify a k-means model in `tidyclust`, simply choose a value of `num_clusters`:

```{r}
kmeans_spec <- k_means(num_clusters = 3)

kmeans_spec
```

There are currently two engines: `stats::kmeans` (default) and 
`ClusterR::KMeans_rcpp`.

It is also possible to change the algorithmic details of the implementation, by
changing the engine and/or using the corresponding arguments from the engine 
functions:

```{r}
kmeans_spec_lloyd <- k_means(num_clusters = 3) |>
  parsnip::set_engine("stats", algorithm = "Lloyd")

kmeans_spec_cr <- k_means(num_clusters = 3) |>
  parsnip::set_engine("ClusterR", initializer = "random")
```

Note that the `stats::kmeans` and the `ClusterR::KMeans_rcpp` implementations have
very different default settings for the algorithmic details, so it is recommended
to be deliberate and explicit in choosing these options. (See the end of this
document for detail on the algorithmic options and defaults.)

## Fitting k-means models

Once specified, a model may be "fit" to a dataset by providing a formula and 
data frame in the same manner as a `tidymodels` model fit.  
Note that unlike in supervised modeling, the formula should not include a 
response variable.

```{r}
kmeans_fit <- kmeans_spec |>
  fit(~ bill_length_mm + bill_depth_mm,
    data = penguins
  )

kmeans_fit |>
  summary()
```

To access only the results produced by the engine - in this case, 
`stats::kmeans` - simply extract the fit from the fitted model object:

```{r}
kmeans_fit$fit
```

`tidyclust` also provides a function, `extract_fit_summary()`, to produce a
list of model summary information in a format that
is consistent across all cluster model specifications and engines

```{r}
kmeans_summary <- kmeans_fit |>
  extract_fit_summary()

kmeans_summary |> str()
```

## Cluster assignments and centers

The primary objective of fitting a clustering model is typically to assign
observations to clusters.  To access these, use `extract_cluster_assignment()` 
function:

```{r}
kmeans_fit |>
  extract_cluster_assignment()
```

Note that this function renames clusters in accordance with the standard `tidyclust`
naming convention and ordering: clusters are named "Cluster_1", "Cluster_2", etc.
and are numbered by the order they appear in the rows of the training dataset.

To reconcile these standardized cluster labels with the engine output, refer
back to the full model fit summary:

```{r}
tibble(
  orig_labels = kmeans_summary$orig_labels,
  standard_labels = kmeans_summary$cluster_assignments
)
```
In this example, we see that the cluster labelled "3" by the `stats::kmeans` 
engine function - a label that is assigned randomly by this implementation -
is the first to appear in the training data, so it is converted to "Cluster_1"
in the standardized labels.

### Centroids

A secondary output of interest is often the characterization of the clusters;
i.e., what data feature trends does each cluster seem to represent?  Most commonly,
clusters are characterized by their mean values in the predictor space, a.k.a.
the centroids.

These can be accessed from the full summary:

```{r}
kmeans_summary$centroids
```

They can also be accessed directly from the fitted model with:

```{r}
kmeans_fit |>
  extract_centroids()
```

```{r}
#| echo: false
centroids <- extract_centroids(kmeans_fit)

small_length <- centroids$.cluster[which.min(centroids$bill_length_mm)]
small_depth <- centroids$.cluster[which.min(centroids$bill_depth_mm)]
other <- setdiff(levels(centroids$.cluster), c(small_length, small_depth))
```


Based on the above output, we might say that `r small_length` is penguins with smaller
bill lengths, `r small_depth` has smaller bill depths, and `r other` is penguins with
large bills in both dimensions.

## Prediction

<!-- Since unsupervised methods are not designed for prediction, the notion of  -->
<!-- predicting the cluster assignment is not always obvious. -->

Since the $k$-means algorithm ultimately assigns training observations to the
cluster with the closest centroid, it is natural to "predict" that test 
observations also belong to the closest centroid cluster.

The `predict()` function behaves as expected, producing cluster assignment 
predictions on new data based on distance to the fitted model centroids.

```{r}
new_penguin <- tibble(
  bill_length_mm = 42,
  bill_depth_mm = 17
)

kmeans_fit |>
  predict(new_penguin)
```
<!-- It is possible to tweak the prediction process by using a different distance -->
<!-- metric to measure closest centroid.  This is not recommended unless the original -->
<!-- model fit also used an alternate distance function in the $k$-means algorithm. -->

<!-- ```{r} -->
<!-- kmeans_fit |> -->
<!--   predict(new_penguin, -->
<!--           dist_fun = dist(., method = "manhattan")) -->
<!-- ``` -->


To attach all predictions to a dataset as a column, use `augment()`:

```{r}
kmeans_fit |>
  augment(penguins)
```

## Metrics

Since clustering is an unsupervised method, with no target/outcome variable, there
is no objective notion of predictive success.

However, many common approaches exist for quantifying the quality of a particular
cluster partition or structure.

### Sum of squared error

One simple metric is the **within cluster sum-of-squared error (WSS)**, which measures the
sum of all distances from observations to their cluster center.  This is 
sometimes scaled with the **total sum-of-squared error (TSS)**, the distance from all
observations to the global centroid; in particular, the ratio **WSS/TSS** is
often computed.  In principle, small values of WSS or of the WSS/TSS ratio suggest
that the observations within clusters are closer (more similar) to each other 
than they are to the other clusters.

The WSS and TSS come "for free" with the model fit summary, or they can be accessed
directly from the model fit:

```{r}
kmeans_summary$sse_within_total_total
kmeans_summary$sse_total

kmeans_fit |> sse_within_total()
kmeans_fit |> sse_total()

kmeans_fit |> sse_ratio()
```

We can also see the within sum-of-squares by cluster, rather than totalled,
with `sse_within()`:

```{r}
kmeans_fit |>
  sse_within()
```

### Silhouette

Another common measure of cluster structure is called the **silhouette**.  The
silhouette of a single observation is proportional to the average distance
from that observation to within-cluster observations minus the average distance
to outside-cluster observations; normalized by the greater of these two average.  
In principle, a large silhouette (close to 1) suggests that an observation is 
more similar to those within its cluster than those outside its cluster.

We can average all silhouettes to get a metric for the full clustering fit. 
Beause the computation of the silhouette depends on the original observation
values, a dataset must also be supplied to the function.

```{r}
kmeans_fit |>
  silhouette_avg(penguins)
```

### Changing distance measures

These metrics all depend on measuring the distance between points and/or
centroids.  By default, ordinary Euclidean distance is used.  However, it is
possible to select a different distance function.  

For sum of squares metrics, the distance function supplied must take two arguments
(i.e., the observation locations and the centroid locations).  For the sihouette
metric, the distance function must find pairwise distances from a single 
matrix (i.e., all pairwise distances between observations).

```{r}
my_dist_1 <- function(x) {
  philentropy::distance(x, method = "manhattan")
}

my_dist_2 <- function(x, y) {
  philentropy::dist_many_many(x, y, method = "manhattan")
}

kmeans_fit |> sse_ratio(dist_fun = my_dist_2)

kmeans_fit |> silhouette_avg(penguins, dist_fun = my_dist_1)
```

For more on using metrics for cluster model selection, see the Tuning vignette.

## Workflows

The workflow structure of `tidymodels` is also usable with `tidyclust` objects.
In the following example, we try two recipes for clustering penguins by bill
dimensions.  In the second recipe, we log-scale both predictors before 
clustering.

```{r}
penguins_recipe_1 <- recipe(~ bill_length_mm + bill_depth_mm,
  data = penguins
)

penguins_recipe_2 <- recipe(~ bill_length_mm + bill_depth_mm,
  data = penguins
) |>
  step_log(all_numeric_predictors())

wflow_1 <- workflow() |>
  add_model(kmeans_spec) |>
  add_recipe(penguins_recipe_1)

wflow_2 <- workflow() |>
  add_model(kmeans_spec) |>
  add_recipe(penguins_recipe_2)

wflow_1 |>
  fit(penguins) |>
  extract_centroids()

wflow_2 |>
  fit(penguins) |>
  extract_centroids()
```


## A brief introduction to the k-means algorithm

*k-means* is a method of *unsupervised* learning that produces a partitioning of
observations into *k* unique clusters.  The goal of *k-means* is to minimize the sum of
squared Euclidian distances between observations in a cluster and the **centroid**,
or geometric mean, of that cluster.

In *k-means* clustering, observed variables (columns) are considered to be locations
on axes in multidimensional space.  For example, in the plot below,
each point represents an observation of one penguin, and the location in 2-dimensional
space is determined by the bill length and bill depth of that penguin.

```{r}
#| echo: false
#| fig-alt: "scatter chart. bill_length_mm along the x-axis, bill_depth_mm along the y-axis. 3 vague cluster appears in the point cloud."
penguins |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point() +
  theme_minimal()
```

A *k-means* cluster assignment is achieved by iterating to convergence from random
initial conditions.  The algorithm typically proceeds as follows:

1. Choose *k* random observations in the dataset.  These locations in space are
declared to be the **initial centroids**.

```{r}
#| echo: false
#| fig-alt: "scatter chart. bill_length_mm along the x-axis, bill_depth_mm along the y-axis. 3 vague cluster appears in the point cloud. 3 random points have been highlighted red, green, and blue."
pens <- penguins |>
  select(bill_length_mm, bill_depth_mm) |>
  drop_na()

init <- sample(seq_len(nrow(pens)), 3)

thing <- kmeans(pens, centers = pens[init, ], iter.max = 1)

centers <- data.frame(thing$centers)

pens |>
  ggplot() +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(
    data = pens[init, ],
    aes(
      x = bill_length_mm, y = bill_depth_mm,
      color = c("Cluster 1", "Cluster 2", "Cluster 3")
    ),
    shape = "o", size = 12, stroke = 1
  ) +
  geom_point(
    data = pens[init, ],
    aes(
      x = bill_length_mm, y = bill_depth_mm,
      color = c("Cluster 1", "Cluster 2", "Cluster 3")
    )
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

2. Assign each observation to the nearest centroid.

```{r}
#| echo: false
#| fig-alt: "scatter chart. bill_length_mm along the x-axis, bill_depth_mm along the y-axis. 3 vague cluster appears in the point cloud. Point are colored according to how close they were to the color points."
closest_center <- philentropy::dist_many_many(as.matrix(pens), as.matrix(pens[init, ]), method = "euclidean") |>
  apply(1, which.min)

pens |>
  ggplot() +
  geom_point(
    data = pens,
    aes(
      x = bill_length_mm, y = bill_depth_mm,
      color = as.factor(closest_center)
    ),
    size = 4
  ) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm)) +
  theme_minimal() +
  theme(legend.position = "none")
```

3. Compute the new centroids of each cluster.

```{r}
#| echo: false
#| fig-alt: "scatter chart. bill_length_mm along the x-axis, bill_depth_mm along the y-axis. 3 vague cluster appears in the point cloud. New highlights are placed in the center of the points of the same color in the previous chart."
centers <- pens |>
  mutate(
    clust = closest_center
  ) |>
  group_by(clust) |>
  summarize_all(mean)

pens |>
  ggplot() +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(
    data = centers,
    aes(
      x = bill_length_mm, y = bill_depth_mm,
      color = c("Cluster 1", "Cluster 2", "Cluster 3")
    ),
    shape = "x", size = 12, stroke = 1
  ) +
  geom_point(
    data = centers,
    aes(
      x = bill_length_mm, y = bill_depth_mm,
      color = c("Cluster 1", "Cluster 2", "Cluster 3")
    )
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


4. Repeat steps 2 and 3 until the centroids do not change.

```{r}
#| echo: false
#| fig-alt: "scatter chart. bill_length_mm along the x-axis, bill_depth_mm along the y-axis. 3 vague cluster appears in the point cloud. All points have been colored of the 3 colors, with an ellipse acound each color."
thing <- kmeans(pens, centers = pens[init, ])

centers <- data.frame(thing$centers)

pens |>
  ggplot() +
  geom_point(
    data = pens,
    aes(
      x = bill_length_mm,
      y = bill_depth_mm,
      color = as.factor(thing$cluster)
    )
  ) +
  geom_mark_ellipse(
    aes(
      x = bill_length_mm,
      y = bill_depth_mm,
      color = as.factor(thing$cluster)
    ),
    expand = unit(1, "mm")
  ) +
  geom_point(
    data = centers,
    aes(
      x = bill_length_mm,
      y = bill_depth_mm,
      color = as.factor(1:3)
    ),
    shape = "x",
    size = 12,
    stroke = 1
  ) +
  geom_point(
    data = pens,
    aes(
      x = bill_length_mm,
      y = bill_depth_mm,
      color = as.factor(thing$cluster)
    )
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


### Iteration of centroids

There is also some variation between implementations on how the update process
takes place.

In the above example, we have shown the common implementation known as the
`Lloyd` or the `Forgy` method.  The update steps are:

1. Assign **all** observations to closest centroid.
2. Recalculate centroids.
3. Repeat until convergence.

One variant on this approach is the `MacQueen` method, which updates centroids
continually:

1. Assign **one** observation to closest centroid.
2. Recalculate centroids.
3. Repeat until all observations have been reassigned once.
4. Repeat until convergence.

A third common variant is the `Hartigan-Wong` method, which assigns observations
based on overall sum of squared errors rather than simply to the closest cluster:

1. Temporarily assign **one** observation to **one** cluster.
2. Recalculate centroid.
3. Find **all** distances from observations to their cluster center (SSE).
4. Repeat for **each** cluster.
5. Permanently assign the observation to the cluster that resulted in the lowest
SSE.
6. Repeat for all observations.
7. Repeat until convergence.


As with many interactive algorithms, the choice between methods is a choice of
complexity versus accuracy.  The `Hartigan-Wong` method generally results in 
more consistent and human-verified clusterings, and it is the default setting
for the `stats::kmeans` implementation of k-means clustering; although all
three algorithms are available as options in this engine.  The `Lloyd/Forgy` 
method is the most simple and ubiquitous; this is the only method available
in the `ClusterR` package implementation.

[Source](https://towardsdatascience.com/three-versions-of-k-means-cf939b65f4ea)


### Initialization of the k-means algorithm

The *k-means* algorithm depends on choosing an initial set of cluster centers.

There are three common methods for selecting initial centers:

1. **Random observations:** In the example above, we have chosen three random 
observations to act as our initial centers.  This is the most commonly used
approach, implemented in the `Forgy`, `Lloyd`, and `MacQueen` methods.

2. **Random partition:** The observations are assigned to a cluster uniformly
at random.  The centroid of each cluster is computed, and these are used
as the initial centers.  This approach is implemented in the `Hartigan-Wong`
method.

3. **k-means++:** Beginning with one random set of the observations, further
observations are sampled via probability-weighted sampling until $k$ clusters
are formed.  The centroids of these clusters are used as the initial centers.
([Further detail here](https://en.wikipedia.org/wiki/K-means%2B%2B))

Because the initial conditions are based on random selection in both approaches,
the k-means algorithm is not determinitistic.  That is, running the clustering
twice on the same data may not result in the same cluster assignments.

It is common to perform the k-means clustering algorithm multiple times, with
different random initial conditions, and combine results at the end.  This option
is controlled by the `nstart` argument of the `stats::kmeans` implementation,
and the `num_init` argument of the `ClusterR::KMeans_rcpp` implementation.


























