---
title: "kmeans_metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{kmeans_metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(celery)
library(tidyverse)
library(tidymodels)

data("penguins", package = "modeldata")
penguins <- penguins %>% drop_na()
```


## Setup

```{r}
kmeans_spec <- k_means(k = 3) %>%
    set_engine_celery("stats") 
  
kmeans_fit <- kmeans_spec %>%
  fit(~ bill_length_mm + bill_depth_mm,
      data = penguins)

kmeans_fit %>% extract_centroids()
kmeans_fit %>% extract_fit_summary()
```


## Metrics

```{r}
kmeans_fit %>% within_cluster_sse()
kmeans_fit %>% tot_wss()
kmeans_fit %>% tot_sse()
kmeans_fit %>% sse_ratio()
```

```{r}
# kmeans_fit %>%
#   augment(penguins)
  
dists <- penguins %>%
  select(contains("bill")) %>%
  dist()

silhouettes(kmeans_fit, dists = dists)

avg_silhouette(kmeans_fit, dists = dists)
```



#### Comparison to auxilary variables

ALERT NAS ARE STUPID

```{r, eval=FALSE}
penguins <- penguins %>%
  drop_na(bill_length_mm, bill_depth_mm) 

penguins %>%
  cbind(extract_cluster_assignment(kmeans_fit)) %>%
  enrichment(.cluster, species)
```



## Cross-Validation

take the full data assignments as "ground truth"


## Tuning

Traditional "elbow" plot:

```{r, eval=FALSE}

results_wss <- c()
results_ratio <- c()

for (k in 2:10) {
  
  kmeans_spec_k <- k_means(k = k) %>%
    set_engine_celery("stats") 
  
  kmeans_fit_k <- kmeans_spec_k %>%
    fit(~ bill_length_mm + bill_depth_mm,
        data = penguins)
  
  
  results_ratio <- c(results_ratio, sse_ratio(kmeans_fit_k))
  results_wss <- c(results_wss, wss(kmeans_fit_k))
}

plot(results_wss)
plot(results_ratio)
```

(Note that the totss aka residual ss is the same no matter what k, so these plots will be the same except the scale of the y-axis.  I hypothesize ratio will be better for cross validation.)


What about by enrichment?


```{r}

results_species <- c()

for (k in 2:10) {
  
  kmeans_spec_k <- k_means(k = k) %>%
    set_engine_celery("stats") 
  
  kmeans_fit_k <- kmeans_spec_k %>%
    fit(~ bill_length_mm + bill_depth_mm,
        data = penguins)
  
  res <- penguins %>%
  cbind(extract_cluster_assignment(kmeans_fit_k)) %>%
  enrichment(.cluster, species)
  
  results_species <- c(results_species, res)

}

plot(results_species)
```

Is there a penalized ChiSq type thing?  Look into this.


## Silhouette


