---
title: "Tuning Cluster Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning Cluster Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE,
  message = FALSE,
  warning = FALSE
)
```

## Setup

Load libraries:

```{r}
#| label: setup
library(tidyclust)
library(tidyverse)
library(tidymodels)
set.seed(838383)
```

Load and clean a dataset:

```{r}
data("penguins", package = "modeldata")

penguins <- penguins |>
  drop_na()
```

## Tuning in unsupervised settings

In *supervised modeling* scenarios, we observe values of a target (or "response")
variable, and we measure the success of our model based on how well it predicts
future response values.  To select hyperparameter values, we **tune** them, trying
many possible values and measuring how well each performs when predicting
target values of test data.

In the *unsupervised modeling* setting of `tidyclust`, there is no such objective
measure of success.  Clustering analyses are typically exploratory rather than
testable.  Nonetheless, the core tuning principle of varying inputs and quantifying
results is still applicable.


## Specify and fit a model

In this example, we will fit a $k$-means cluster model to the `palmerpenguins`
dataset, using only the bill length and bill depth of penguins as predictors.

(Please refer to the k-means vignette for an in-depth discussion of this model
specification.)

Our goal will be to select an appropriate number of clusters for the model 
based on metrics.

First, we set up cross-validation samples for our data:

```{r}
penguins_cv <- vfold_cv(penguins, v = 5)
```

Next, we specify our model with a tuning parameter, make a workflow,
and establish a range of possible values of `num_clusters` to try:

```{r}
kmeans_spec <- k_means(num_clusters = tune())

penguins_rec <- recipe(~ bill_length_mm + bill_depth_mm,
  data = penguins
)

kmeans_wflow <- workflow(penguins_rec, kmeans_spec)

clust_num_grid <- grid_regular(num_clusters(),
  levels = 10
)

clust_num_grid
```

Then, we can use `tune_cluster()` to compute metrics on each cross-validation
split, for each possible choice of number of clusters.

```{r}
res <- tune_cluster(
  kmeans_wflow,
  resamples = penguins_cv,
  grid = clust_num_grid,
  control = control_grid(save_pred = TRUE, extract = identity),
  metrics = cluster_metric_set(sse_within_total, sse_total, sse_ratio)
)

res
```

```{r}
res_metrics <- res |> collect_metrics()
res_metrics
```

### Choosing hyperparameters

In supervised learning, we would choose the model with the best value of a
target metric.  However, clustering models in general have no such local maxima 
or minima.  With more clusters in the model, we would always expect the within
sum-of-squares to be smaller.

A common approach to choosing a number of clusters is to look for an "elbow",
or notable bend, in the plot of WSS/TSS ratio by cluster number:

```{r}
#| fig-alt: "Connected line chart. Number of clusters along the x-axis, mean WSS/TSS ratio, over 5 folds along the y-axis. X-axis ranges from 1 to 10, with 1 having an value of 1, 2 having a value of 0.35, and the rest haver an ever decreasing value."
res_metrics |>
  filter(.metric == "sse_ratio") |>
  ggplot(aes(x = num_clusters, y = mean)) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  ylab("mean WSS/TSS ratio, over 5 folds") +
  xlab("Number of clusters") +
  scale_x_continuous(breaks = 1:10)
```

At each increase in the number of clusters, the WSS/TSS ratio decreases, with the
amount of decrease getting smaller as the number of clusters grows.  We might
argue that the drop from two clusters to three, or from three to four, is a bit
more extreme than the subsequent drops, so we should probably choose three or
four clusters.



<!-- ## Comparison to auxilary variables -->

<!-- It is sometimes of interest to compare to external variables -->

<!-- ```{r, eval=FALSE} -->
<!-- kmeans_fit |> -->
<!--   augment(penguins) |> -->
<!--   enrichment(.pred_cluster, species) -->
<!-- ``` -->

